Doublet splitting of fusion alpha-particle driven ion cyclotron emission

### 2022-07-28 manuscript.tex
final proof

\address{UKAEA-CCFE, Culham Science Centre, Abingdon, OX14 3DB, UK}
\ead{james.cook@ukaea.uk}
\vspace{10pt}
\begin{indented}
\item[]Jan 2022
\end{indented}

@@

The foremost explanation of the mechanism responsible for ICE is the
magnetoacoustic cyclotron instability (MCI) \cite{Belikov1976}, which arises from
the free energy present in energetic ions with inverted distributions in
velocity space. Theoretical understanding of ICE is based on analytical
solutions to the linearised Maxwell-Vlasov system of equations \cite{Stix} where
the energetic ion species is modelled as ring-beam for which $f(v_\parallel,
v_\perp) \propto \exp(-(v_\parallel-u_\parallel)^2/ v_{th}^2)\delta(v_\perp -
u_\bot)$ \cite{Dendy1994, McClements1996}. Here $u_\parallel$, $u_\perp$ and
$v_{th}$ are the beam velocity (the signed drift speed along the magnetic
field), ring speed (the speed of the shifted centre of the inverted population
perpendicular to the magnetic field)
and thermal velocity, respectively. Subsequent studies augmented this
understanding via nonlinear fully-kinetic \cite{Cook2013,Chapman2017} and
hybrid \cite{Carbajal2014,Reman2019} particle-in-cell (PIC) simulations in one
spatial dimension with three vector components for velocities and electromagnetic
field (i.e. 1D3V), and recently 2D3V hybrid-PIC simulations \cite{Carbajal2021}. These
computational approaches operate under similar simplifications as the analytical
treatments in that the plasma is initially spatially homogeneous and periodic
in the spatial domain.
Linearised fluid approaches that incorporate toroidal geometry and
inhomogeneous plasma and magnetic fields have shed light of the eigenmode
structure of compressional Alfv{\'e}n eigenmodes, which as with ICE are also fast
Alfv{\'e}n modes driven by energetic ions, in spherical
tokamaks \cite{Sharapov2014, Gorelenkov2016} and conventional aspect ratio
tokamaks \cite{Kolesnichenko2000, Fulop2000, Heidbrink2006}.

@@

Solutions to the linearised Maxwell-Vlasov system of equations are found in
perpendicular and parallel wavenumber space, $(k_\bot, k_\parallel)$, for a
ring-beam distribution of energetic minority
alpha-particles, Maxwellian electrons and background thermal majority
deuteron ions. The ring-beam distribution function, Eq. 1, is a separable
function of the beam in $v_\parallel$ given by Eq. 2 and the
ring in $v_\perp$ given by Eq. 3.
In the context of the fully kinetic Maxwell-Vlasov linear theory,
which rests upon the assumption of
an infinite homogeneous plasma and a straight uniform magnetic field,
this distribution function is parameterised by the drift
velocity parallel to the magnetic field $u_\parallel$, the speed of the
Maxwellian distribution perpendicular to the magnetic
field $u_{\perp}$, and the isotropric thermal velocity $v_{th}$. The
thermal spread of the alpha-particles' distribution function 
$v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The number density
ratio of alpha-particles to the majority fuel deuterons is $1.5\times10^{-4}$.

@@

Calculations are made for alpha-particles with ring-beams of two
values of pitch-angle cosine: $u_\parallel=-0.646 u_0 \approx
-\cos(\arcsin(\sqrt{R_0 \sin(\psi_0)/R_{ICE}})) u_0 \simeq -1.1 V_A$, which is
associated with the ion-cyclotron emission observed on JET Pulse No. 26148 for
which $R_0=3.46$ m and
$\psi_0=55.2^\circ$ are the major radius and pitch angle at birth, and
$R_{ICE}=4$ m is the major radius of the emission region; see
Figs. 3 and 14 and associated text in Ref. \cite{Cottrell1993}; and
$(u_\parallel=0 u_0, u_\bot=u_0)$ representing a ring distribution with
no parallel drift, see
Fig. 1(a) of Ref \cite{Cook2013}. Here, $V_A$ is the Alfv{\'en} speed and $u_0=\sqrt{u_\parallel^2 + u_\perp^2}$ is
the speed of an alpha-particle with an energy of 3.6 MeV.

@@

\begin{align}
    0 &= ||\vec{\vec{\epsilon}} + \left(\vec{k} \otimes \vec{k} - \vec{\vec{I}}(\vec{k}\cdot\vec{k})\right) \frac{c^2}{\omega^2}||,\label{eq:dettensor}\\
    \vec{\vec{\epsilon}} &= \vec{\vec{1}} + \sum_s \frac{\Pi_s^2}{\omega}\sum_{n=-\infty}^{\infty}
    \int_0^{\infty}
    \int_{-\infty}^{\infty}\frac{\vec{\vec{S}}_{s,n} 2\pi v_\perp dv_\perp dv_\parallel}{\omega - k_\parallel v_\parallel - n \Omega_s}
    \label{eq:dielectrictensor}
\end{align}

@@

Integrals over perpendicular velocity required for calculating the
dielectric tensor for Maxwellian distribution functions are given in terms of
modified Bessel functions of the first kind \cite{Stix} for both electrons and
deuterons i.e. when $u_\perp =0$. The same perpendicular integrals over the
alpha-particles' ring distribution given by Eq. 3 are calculated numerically via
Gauss-Kronrod \cite{Kronrod1965} quadrature using the GaussGK.jl \cite{QuadGK}
library. An absolute and relative tolerance of $0$ and $10^{-8}$, respectively,
is used to confirm convergence of the integral values.
Since all the parallel distribution functions for all species are Maxwellians,
the parallel integrals can be calculated from the well known plasma dispersion
function\cite{Fried1961}, which is laid out in the appendix.

@@

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig2.pdf}
    \caption{(colour online) Shading indicates the growth rate of fast
    Alfv{\'e}n waves excited by a ring of fusion born alpha-particles, as a
    function of parallel wavenumber on the ordinate and perpendicular wavenumber
    on the abscissa. No solution is found where shading is absent. The
    alpha-particle distribution function is given by Eq. 1 with
    $u_{\parallel} = 0$. Consecutive straight traces indicate the propagation
    angle with respect to the magnetic field: the horizontal line is exactly
    perpendicular; solid traces fan out at intervals of five degrees; and dashed
    traces indicate intervals of one degree. Curved grey traces indicate the
    location in $(k_\perp, k_\parallel)$ of the cyclotron harmonics (i.e. the
    real component of the complex $\omega$ solutions), which are annotated along
    the left hand side and upper edge of the figure. Frequencies are in units of
    the alpha-particle cyclotron frequency, and wavenumbers are in
    alpha-particle cyclotron frequencies per Alfv{\'e}n speed. Note the up-down
    symmetry due to the lack of parallel drift in the energetic alpha
    particles.}
    \label{fig:2D_imag_zeropitch}
\end{figure}

@@

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions
since they arise from contributions from a range of propagation angles.
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes
more clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$k_\bot \geq 7\Omega_i/V_A$ is dominated by broader bands of instability
near to Doppler shifted cyclotron harmonics, which have
an intricate structure in wavenumber space. Modes numbering 1-5 and
14 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 13 \Omega_i/V_A$, i.e. a propagation
angle of $\sim 93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling modes and
propagation angles that are close to perpendicular.

Figure \ref{fig:2D_real_zeropitch}
shows the smoothed real frequency part of the solution, as depicted
by contours on Figs. \ref{fig:2D_imag_zeropitch} and
\ref{fig:2D_imag_physicalpitch}, for the case of
zero pitch angle cosine and is normalised by
the high frequency expression for the fast Alfv{\'e}n wave set out in Eq.
25 of Ref. \cite{Dendy1994} (also Eq. 11 of Ref. \cite{McClements1996}),

\begin{equation}
  \omega^2 = \frac{V_A^2}{2}\left[
    k^2 + k_\parallel^2 + k^2k_\parallel^2 \frac{V_A^2}{\Omega_i^2}
  + \sqrt{
    \left(k^2 + k_\parallel^2 + k^2k_\parallel^2 \frac{V_A^2}{\Omega_i^2}\right)^2
    - 4k^2 k_\parallel^2
}
  \right],
\label{eqn:fastalfven}
\end{equation}

@@

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig5.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except that
    the shading indicates the real part of the smoothed solution normalised by
    the high frequency expression for the fast Alfv{\'e}n wave set out in
    Eq. \label{eqn:fastalfven} (Eq. 25 of Ref. \cite{Dendy1994},
    which is also Eq. 11 in Ref. \cite{McClements1996}).
    The differences in real frequency for the two sets of
    calculations, a pitch-angle cosine of $-0.646$ and zero, are
    negligible.}
    \label{fig:2D_real_zeropitch}
\end{figure}

@@

This paper presents the linear Maxwell-Vlasov dispersion relation of a plasma
consisting of thermal electrons, thermal deuteron fuel ions and a minority
energetic ring-beam of fusion born alpha-particles. These high resolution
$(k_\perp, k_\parallel)$ calculations revisit ICE observations from JET Pulse
No. 26148 \cite{Cottrell1993,Dendy1995} and reproduce several important features:
the step-change in signal strength from $\omega \leq 7\Omega_i$
to $\omega > 7\Omega_i$; the
doublet splitting of peaks at $\omega \approx 10\Omega_i$ in Fig.
\ref{fig:2D_F12_physicalpitch}, with origins visible in the fine $(k_\bot,
k_\parallel)$ structure in Fig. \ref{fig:2D_imag_physicalpitch}. The Doppler shift
for large $k_\parallel$ becomes of the order of the alpha-particle cyclotron
frequency making it difficult to determine which parts of the signal come
from which cyclotron harmonic numbers, which has important implications for
determining the major radius of origin of ICE.
Further effort is required to account for
the extra doublets visible in the continuous trace associated with JET Pulse No.
26148 in Fig. 2 of Ref. \cite{Cottrell1993}.

>>>>

\address{UKAEA-CCFE, Culham Science Centre, Abingdon, OX14 3DB, UK}
\ead{james.cook@ukaea.uk}
\vspace{10pt}
\begin{indented}
\item[]Jul 2022
\end{indented}

@@

The foremost explanation of the mechanism responsible for ICE is the
magnetoacoustic cyclotron instability (MCI) \cite{Belikov1976}, which arises from
the free energy present in energetic ions with inverted distributions in
velocity space. Theoretical understanding of ICE is based on analytical
solutions to the linearised Maxwell-Vlasov system of equations \cite{Stix} where
the energetic ion species is modelled as ring-beam for which $f(v_\parallel,
v_\perp) \propto \exp(-(v_\parallel-u_\parallel)^2/ v_{th}^2)\delta(v_\perp -
u_\bot)$ \cite{Dendy1994, McClements1996}. Here $u_\parallel$, $u_\perp$ and
$v_{th}$ are the beam velocity (the signed drift speed along the magnetic
field), ring speed (the speed of the shifted centre of the inverted population
perpendicular to the magnetic field)
and thermal velocity, respectively. Subsequent studies augmented this
understanding via nonlinear fully-kinetic \cite{Cook2013,Chapman2017} and
hybrid \cite{Carbajal2014,Reman2019} particle-in-cell (PIC) simulations in one
spatial dimension with three vector components for velocities and electromagnetic
field (i.e. 1D3V), and recently 2D3V hybrid-PIC simulations \cite{Carbajal2021}. These
computational approaches operate under similar simplifications as the analytical
treatments in that the plasma is initially spatially homogeneous and periodic
in the spatial domain.
Linearised fluid approaches that incorporate toroidal geometry and
inhomogeneous plasma and magnetic fields have shed light on the eigenmode
structure of compressional Alfv{\'e}n eigenmodes, which as with ICE are also fast
Alfv{\'e}n modes driven by energetic ions, in spherical
tokamaks \cite{Sharapov2014, Gorelenkov2016} and conventional aspect ratio
tokamaks \cite{Kolesnichenko2000, Fulop2000, Heidbrink2006}.

@@

Solutions to the linearised Maxwell-Vlasov system of equations are found in
perpendicular and parallel wavenumber space, $(k_\bot, k_\parallel)$, for a
ring-beam distribution of energetic minority
alpha-particles, Maxwellian electrons and Maxwellian majority
deuteron ions. The ring-beam distribution function, Eq. 1, is a separable
function of the beam in $v_\parallel$ given by Eq. 2 and the
ring in $v_\perp$ given by Eq. 3.
In the context of the fully kinetic Maxwell-Vlasov linear theory,
which rests upon the assumption of
an infinite homogeneous plasma and a straight uniform magnetic field,
this distribution function is parameterised by the drift
velocity parallel to the magnetic field $u_\parallel$, the shift of the
Maxwellian distribution perpendicular to the magnetic
field $u_{\perp}$, and the isotropric thermal velocity $v_{th}$. The
thermal spread of the alpha-particles' distribution function 
$v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The number density
ratio of alpha-particles to the majority fuel deuterons is $1.5\times10^{-4}$.

@@

Calculations are made for alpha-particles with ring-beams of two
values of pitch-angle cosine: $u_\parallel=-0.646 u_0 \approx
-\cos(\arcsin(\sqrt{R_0 \sin(\psi_0)/R_{ICE}})) u_0 \simeq -1.1 V_A$, which is
associated with the ion-cyclotron emission observed on JET Pulse No. 26148 for
which $R_0=3.46$ m and
$\psi_0=55.2^\circ$ are the major radius and pitch-angle at birth, and
$R_{ICE}=4$ m is the major radius of the emission region: see
Figs. 3 and 14 and associated text in Ref. \cite{Cottrell1993}; and
$(u_\parallel=0 u_0, u_\bot=u_0)$ representing a ring distribution with
no parallel drift, see
Fig. 1(a) of Ref \cite{Cook2013}. Here, $V_A$ is the Alfv{\'en} speed and $u_0=\sqrt{u_\parallel^2 + u_\perp^2}$ is
the speed of an alpha-particle with an energy of 3.6 MeV.

@@

\begin{align}
    0 &= |\vec{\vec{\epsilon}} + \left(\vec{k} \otimes \vec{k} - \vec{\vec{I}}(\vec{k}\cdot\vec{k})\right) \frac{c^2}{\omega^2}|,\label{eq:dettensor}\\
    \vec{\vec{\epsilon}} &= \vec{\vec{1}} + \sum_s \frac{\Pi_s^2}{\omega}\sum_{n=-\infty}^{\infty}
    \int_0^{\infty}
    \int_{-\infty}^{\infty}\frac{\vec{\vec{S}}_{s,n} 2\pi v_\perp dv_\perp dv_\parallel}{\omega - k_\parallel v_\parallel - n \Omega_s}
    \label{eq:dielectrictensor}
\end{align}

@@

Integrals over perpendicular velocity required for calculating the
dielectric tensor for Maxwellian distribution functions are given in terms of
modified Bessel functions of the first kind \cite{Stix} for both electrons and
deuterons i.e. when $u_\perp =0$. The same perpendicular integrals over the
alpha-particles' ring distribution given by Eq. 3 are calculated numerically via
Gauss-Kronrod \cite{Kronrod1965} quadrature using the GaussGK.jl \cite{QuadGK}
library. An absolute and relative tolerance of $0$ and $10^{-8}$, respectively,
is used to confirm convergence of the integral values.
Since all the parallel distribution functions for all species are Maxwellians,
the parallel integrals can be calculated from the well-known plasma dispersion
function\cite{Fried1961}, which is laid out in the appendix.

@@

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig2.pdf}
    \caption{(colour online) Shading indicates the growth rate of fast
    Alfv{\'e}n waves excited by a ring of fusion born alpha-particles, as a
    function of parallel wavenumber on the ordinate and perpendicular wavenumber
    on the abscissa. No solution is found where shading is absent. The
    alpha-particle distribution function is given by Eq. 1 with
    $u_{\parallel} = 0$. Consecutive straight traces indicate the propagation
    angle with respect to the magnetic field: the horizontal line is exactly
    perpendicular; solid traces fan out at intervals of five degrees; and dashed
    traces indicate intervals of one degree. Curved grey traces indicate the
    location in $(k_\perp, k_\parallel)$ of the cyclotron harmonics (i.e. the
    real component of the complex $\omega$ solutions), which are annotated along
    the left-hand side and upper edge of the figure. Frequencies are in units of
    the alpha-particle cyclotron frequency, and wavenumbers are in
    alpha-particle cyclotron frequencies per Alfv{\'e}n speed. Note the up-down
    symmetry due to the lack of parallel drift in the energetic alpha
    particles.}
    \label{fig:2D_imag_zeropitch}
\end{figure}

@@

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$.
Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resemble the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonic is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot
be recreated with a 1D calculation for these distribution functions
since they arise from contributions from a range of propagation angles.
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes
more clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$k_\bot \geq 7\Omega_i/V_A$ is dominated by broader bands of instability
near to Doppler shifted cyclotron harmonics, which have
an intricate structure in wavenumber space. Modes numbering 1-5 and
14 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 13 \Omega_i/V_A$, i.e. a propagation
angle of $\sim 93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling modes and
propagation angles that are close to perpendicular.

Figure \ref{fig:2D_real_zeropitch}
shows the smoothed real frequency part of the solution, as depicted
by contours on Figs. \ref{fig:2D_imag_zeropitch} and
\ref{fig:2D_imag_physicalpitch}, for the case of
zero pitch-angle cosine and is normalised by
the high frequency expression for the fast Alfv{\'e}n wave set out in Eq.
25 of Ref. \cite{Dendy1994} (also Eq. 11 of Ref. \cite{McClements1996}),

\begin{equation}
  \omega^2 = \frac{V_A^2}{2}\left[
    k^2 + k_\parallel^2 + k^2k_\parallel^2 \frac{V_A^2}{\Omega_i^2}
  + \sqrt{
    \left(k^2 + k_\parallel^2 + k^2k_\parallel^2 \frac{V_A^2}{\Omega_i^2}\right)^2
    - 4k^2 k_\parallel^2
}
  \right]\label{eqn:fastalfven},
\end{equation}

@@

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig5.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except that
    the shading indicates the real part of the smoothed solution normalised by
    the high frequency expression for the fast Alfv{\'e}n wave set out in
    Eq. \ref{eqn:fastalfven} (Eq. 25 of Ref. \cite{Dendy1994},
    which is also Eq. 11 in Ref. \cite{McClements1996}).
    The differences in real frequency for the two sets of
    calculations, a pitch-angle cosine of $-0.646$ and zero, are
    negligible.}
    \label{fig:2D_real_zeropitch}
\end{figure}

@@

This paper presents the linear Maxwell-Vlasov dispersion relation of a plasma
consisting of thermal electrons, thermal deuteron fuel ions and a minority
energetic ring-beam of fusion born alpha-particles. These high resolution
$(k_\perp, k_\parallel)$ calculations revisit ICE observations from JET Pulse
No. 26148 \cite{Cottrell1993,Dendy1995} and reproduce several important features:
the step-change in signal strength from $\omega \leq 7\Omega_i$
to $\omega > 7\Omega_i$; the
doublet splitting of peaks at $\omega \approx 10\Omega_i$ in Fig.
\ref{fig:2D_F12_physicalpitch}, with origins visible in the fine $(k_\bot,
k_\parallel)$ structure in Fig. \ref{fig:2D_imag_physicalpitch}. The Doppler shift
for large $k_\parallel$ is comparable in magnitude to the alpha-particle cyclotron
frequency making it difficult to determine which parts of the signal come
from which cyclotron harmonic numbers. This has important implications for
determining the major radius of origin of ICE.
Further effort is required to account for
the extra doublets visible in the continuous trace associated with JET Pulse No.
26148 in Fig. 2 of Ref. \cite{Cottrell1993}.
